Hydra battles in Coq

### 2020-09-14 doc/chapter-ordinals.tex
minor modifications

The proof of termination of all hydra battles presented in~\cite{KP82} is based
on \emph{ordinal numbers}.
From a mathematical point of view, an ordinal is a representant of an equivalence class for isomorphims of strict, total and well-founded ordered sets.

More intuitively, let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a sequence of ordinal numbers, as an extension of the sequence of natural numbers.

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (displayed in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\subsection{Instances of notation systems}

\subsubsection{The ordinal \texorpdfstring{$\omega$}{omega}}

@@

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.Prelude.Finite_ordinals.html}{Prelude.Finite\_ordinals}}

@@

We have now to prove that there exists at most one  proof of \texttt{Nat.ltb x0 (S n)}. This is not obvious, but  a consequence of the following lemma of library 
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Logic.Eqdep_dec.html}{Coq.Logic.Eqdep\_dec}.

>>>>

The proof of termination of all hydra battles presented in~\cite{KP82} is based
on \emph{ordinal numbers}.
From a mathematical point of view, an ordinal is a representant of an equivalence class for isomorphims of  totally ordered well-founded sets.

More intuitively, let us have a look at Figure~\ref{fig:ordinal-sequence}. It presents a small sequence of ordinal numbers, which extends the sequence of natural numbers.

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (displayed in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\section{Instances of notation systems}

\subsection{The ordinal \texorpdfstring{$\omega$}{omega}}

@@

\subsection{The ordinal \texorpdfstring{$\omega^2$}{omega^2}}

The ordinal $\omega^2$ (also called $\varphi_0(2)$, see Chap.~\ref{chap:schutte}), is represented by the lexicographically ordered cartesian product $\mathbb{N}\times\mathbb{N}$.

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.ON_Omega2.html}{ON\_Omega2}}

\begin{Coqsrc}
Definition t := (nat * nat)%type.

Definition lt : relation t := lexico Peano.lt Peano.lt.
Definition le := clos_refl _ lt.

Infix "o<" := lt : o2_scope.
Infix "o<=" := le : o2_scope.
\end{Coqsrc}

It is easy to represent the finite ordinals, successors and limits.

\begin{Coqsrc}
Definition zero: t := (0,0).

Definition fin (n:nat) : t := (0, n).

Notation "'F'" := fin : o2_scope.

Coercion fin : nat >-> t.

Definition succ (alpha : t) := (fst alpha, S (snd alpha)).

Notation "'omega'" := (1,0) : o2_scope.
\end{Coqsrc}

@@

\vspace{4pt}
\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.ON_Finite.html}{OrdinalNotations.ON\_Finite}}

@@

We have now to prove that there exists at most one  proof of (\texttt{Nat.ltb x0 (S n)}). This is not obvious, but  a consequence of the following lemma of library 
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Logic.Eqdep_dec.html}{Coq.Logic.Eqdep\_dec}.
